Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-12 analysis/paper/paper.Rmd
updates for dicussion

@@

Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, trade ornaments specifically, after the presence of the European that hints the emergence of social inequality. We examine the degree of the ceramic standardization to measure ceramic specialization that is usually associate with social differentiation, such as the elite control [@Costin2001; @Junker1999]. The results confirm that the differences in pottery shapes can be detected using EFA method, indicating the foreign influences on the indigenous society- the Europeans in the 17th century and the Chinese in the 19th century. The result of MANOVA shows there is significant difference in shapes between pre-European and Post-European period, and between the pre-European and the Chinese period. The average shape presents round body with wider rim and neck before the European contact, and turns to oval-shaped body with narrower rim and neck after the European presence and even more obvious during the Chinese contact period. In general, pottery shapes become shorter in hight over time that leads to oval-shaped body. For the standardization of pottery shape, the result of CV test indicates the significant differences between pre-European period and post-European period (P-value < 0.05), and between Chinese contact and either pre-European or post-European (P-value < 0.01). This suggests pottery shape becomes more homogeneous and standardized after foreign contacts with the European colonizer and even more with the Chinese immigrants. The compositional analysis shows clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting the raw material sources remains unchanged. More standardized shape in the Chinese period might indicate the appearance of specialized group. To figure out whether this change is related to the function of pot, we made an attempt to conduct lipid analysis for potsherds samples to identify the source of residue absorbed into the fabric. However, we did not obtain useful results due to a low lipid yield, which might result from the thin and less porous fabric where not enough room for absorbing food residue [citation].

The results support part of our hypothesis that foreign influences had impacts on the emergence of social inequality in local indigenous society reflected by the degree of pottery standardization, however, stronger in the Chinese period instead of the European period originally thought to be. The hypothesis of emergence of social inequality based on the monopolies for long-distance trade brought by the Europeans was not fully supported. The nature and context of the interaction with the foreign presences might give an insight into the different degree of standardized pottery. Compared to other region in Taiwan, the European colonial control was weak in Yilan due to the relatively isolated location by mountains and the economic consideration of the Spanish and the Dutch who prefer northern Taiwan as trading base. Yilan could be seen as a place experienced indirect influences by the European trade network and their colonial activities in a pericolonial context [@Acabado2017]. On the contrary, the interaction between indigenous people and the Chinese in the 19th century was more intense and direct since those Chinese immigrants settled in Yilan and lived closely with indigenous societies. This direct influence is reflected by the archaeological evidence of large amount of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009]. Also, the finding of adoption of coffin in mortuary practice that viewed as an ethnic symbol for Chinese was observed [@Chen2007].

More standardized pottery shape without controlling the clay paste or changes in technology suggests the shape as a preferred characteristic in pottery production. In addition to the appearance of specialized groups, another explanation could be related to an act of resistance to the Chinese immigrant by emphasizing the traditional pottery production practice, of which shape is a visible trait for. @Raux2015

@@

We would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the pottery used in this study. We thank Dr. Wen-Shan Chen in the Department of Geosciences, National Taiwan University for his invaluable guidance for petrographic analysis in his lab. This research used statistical consulting resources provided by the Center for Statistics and the Social Sciences, University of Washington.

>>>>

Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, trade ornaments specifically, after the presence of the European that hints the emergence of social inequality. We examine the degree of the ceramic standardization to measure ceramic specialization that is usually associate with social differentiation, such as the elite control [@Costin2001; @Junker1999]. The results confirm that the differences in pottery shapes can be detected using EFA method, indicating the foreign influences on the indigenous society- the Europeans in the 17th century and the Chinese in the 19th century. The result of MANOVA shows there is significant difference in shapes between pre-European and Post-European period, and between the pre-European and the Chinese period. The average shape presents round body with wider rim and neck before the European contact, and turns to oval-shaped body with narrower rim and neck after the European presence and even more obvious during the Chinese contact period. In general, pottery shapes become shorter in hight over time that leads to oval-shaped body. For the standardization of pottery shape, the result of CV test indicates the significant differences between pre-European period and post-European period (P-value < 0.05), and between Chinese contact and either pre-European or post-European (P-value < 0.01). This suggests pottery shape becomes more homogeneous and standardized after foreign contacts with the European colonizer and even more with the Chinese immigrants. The compositional analysis shows clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting the raw material sources remains unchanged. More standardized shape in the Chinese period might indicate the appearance of specialized group. To figure out whether this change is related to the function of pot, we made an attempt to conduct lipid analysis for potsherds samples to identify the source of residue absorbed into the fabric. However, we did not obtain useful results due to a low lipid yield, which might result from the thin and less porous fabric that offers limited spaces to protect organic molecules from microbiological degradation [@Evershed2008, pp. 909].

The results support part of our hypothesis that foreign influences had impacts on the emergence of social inequality in local indigenous society reflected by the degree of pottery standardization, however, stronger in the Chinese period instead of the European period originally thought to be. The hypothesis of emergence of social inequality based on the monopolies for long-distance trade brought by the Europeans was not fully supported. The nature and context of the interaction with the foreign presences might give an insight into the different degree of standardized pottery. Compared to other region in Taiwan, the European colonial control was weak in Yilan due to the relatively isolated location by mountains and the economic consideration of the Spanish and the Dutch who prefer northern Taiwan as trading base. Yilan could be seen as a place experienced indirect influences by the European trade network and their colonial activities in a pericolonial context [cf. @Acabado2017]. On the contrary, the interaction between indigenous people and the Chinese in the 19th century was more intense and direct since those Chinese immigrants settled in Yilan and lived closely with indigenous societies. This direct influence is reflected by the archaeological evidence of large amount of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009]. Also, the finding of adoption of coffin in mortuary practice that viewed as an ethnic symbol for Chinese was observed [@Chen2007].

More standardized pottery shape without controlling the clay paste or technology suggests the shape as a preferred characteristic in pottery production during the Chinese period. However, the standardization is not reflected by the metric attributes, specifically the rim, neck, and body thickness. There is higher variation in thickness attributes compared to diameter attributes, indicating shape is the only attribute that presents homogeneous. In addition to the emergence of specialized groups indicative of changes in social organization, another explanation is that producing homogeneous pottery is an act of resistance to the Chinese immigrant by emphasizing the practice of traditional pottery production. Among the pottery attributes, shape is the most  obvious and visible trait, which can be used as an expression of social boundary or group identity [cf. @Raux2015].

@@

@@

We would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the pottery used in this study and providing the shape images. We thank Dr. Wen-Shan Chen in the Department of Geosciences, National Taiwan University for his invaluable guidance of petrographic analysis at his lab. This research used statistical consulting resources provided by the Center for Statistics and the Social Sciences, University of Washington.
